---
title: "Midpoint Deliverable"
author: "Emily Yu, William Huang, Mekides Demile, Quynh Doan"
output: html_document
---

```{r setup, include=FALSE}
# switch to option of echo = FALSE for the entire file, so
# no codes are shown in the report
knitr::opts_chunk$set(echo = FALSE)
source("scripts/Summary.R")
source("scripts/Table.R")
source("scripts/Chart 1.R")
source("scripts/Chart 2.R")
source("scripts/Chart 3.R")
# library
library(knitr)
# read in data
regular_season_df <- read.csv(
  "./data/2018-2019-NBA-Player-Stats-NBAstuffer-regularseasons.csv",
  stringsAsFactors = FALSE)
playoff_df <- read.csv(
  "./data/2018-2019-NBA-Player-Stats-NBAstuffer-playoffs.csv",
  stringsAsFactors = FALSE)
```

## Introduction
Ａ lot of sports were put on hold in light of health safety concerns during the
COVID period. We are interested in using the data of the latest NBA game season
to look ahead and evaluate each team and each individual player’s performance.
We will be using the NBA player stats dataset for the 2018-2019 game season as 
the 2019-2020 season was suspended due to the pandemic.
[This](https://www.nbastuffer.com/2018-2019-nba-player-stats/) is the link to
the website of the used datasets.

## Summary Information
```{r summary, message=FALSE, warning=FALSE}
# render the summary information which include 5 values
# function returning values that summarize the dataset
info <- get_summary_info(regular_season_df)
```

There were **`r info$player_num`** players and **`r info$team_num`** teams
participated in the latest 2018-2019 NBA season. Several players' performaces
were particularly outstanding, including **`r info$best_player`** who had
played only `r info$best_player_game_played` games in this season and had the
highest VI value, a measure of player's ability to produce points, assists, and
rebounds, among all players, and **`r info$highest_point_player`** who got the
highest points per game in this season. Audiences and fans are specially looking
forward to these talented players' performances in the future.


## Summary Table
To better evaluate each individual player's performance, it is important the
understand the difference between each position in the courts. Each position
focused on different strategies and contributes differently to the game and
the earned scores.

```{r kable_table, message=FALSE, warning=FALSE}
# rendering table
# function returning aggregated information of each position's performance data
position_differences <- position_difference(regular_season_df)
# well-formatted column names are handled in the function
kable(position_differences, caption = "Average Performance Data by Position")
```

### Important Insights from the Table
The aggregated table shows the summarized information grouped by the players'
positions in the courts. The table is sorted so that it is showing the position
with the highest average scoring to the position with the lowest. This table reveals
that each position contributes to the game in different ways as some positions focus
on shooting to get points and some positions focus on defending. These values do
not define the importance of a particular position, instead, the values help us
understand the different nature of each position.

## Bar Chart
VI, Versatility Index, is a metric that measures a player’s ability to produce in
points, assists, and rebounds. To better evaluate each team's ability and the
chance of winning in the future seasons, a bar chart showing the average VI values
of all players in each team can be really helpful. The visualization can help us
understand each team's ability as well as finding out the team with the highest
chance of winning in the future.

```{r bar_chart, message=FALSE, warning=FALSE}
# rendering the bar chart
# function returning a visualization of each team's average player VI value
team_VI_value <- get_bar_chart(regular_season_df)
plot(team_VI_value)
```

### Important Insights from the Bar Chart
According to the bar chart grouped by the team, the team with the highest average
VI values is BOS, meaning that it has a higher chance of winning in the future
based on the performance of this latest game season. It is obvious that there are
factors that may affect the team's performance in future seasons, such as 
improvement and injury, so we are just trying to make the best prediction and find
the team with the highest chance of winning in the future entirely based on the
existed data. There can be star players as well as rookies in each team, so we
define each team's ability by using the average of players' VI values.

## Scatter plot
To better evaluate each team's performance, a scatter plot with two different
colors representing the VI values in the regular season and the playoffs can be
useful. This visualization can help us find out if there are teams trying to
preserve their scoring, save their strength, or perhaps let rookies have more
experiences of participating in the real NBA games while doing the playoffs.

```{r scatter_plot, message=FALSE, warning=FALSE}
# rendering scatterplot
# function returning visualization of each team's player VI which is colored
# to represent different type of games, the regular season versus playoffs
regular_vs_playoff <- get_scatter_plot(regular_season_df, playoff_df)
plot(regular_vs_playoff)
```

### Important Insights from the Scatter Plot
Based on the scatter plot, we can notice that for teams that participate in both
regular season and the playoffs have similar VI value distributions, meaning that
there are no teams with obvious intent to save their strength while doing the 
playoffs. This is a little bit different from our initial expectation, but the
scatter plot reveals that all teams try their best in both types of games and our
assumption is wrong.

## Pie Chart
To understand more about players in each team, a grouped pie chart can help us
understand the players' experience distribution of each team. We define a player's
experience by using the value of GP, games played. We can understand the
distribution of players who participate in numerous games and the percentage of
players who are new players with less experiences in each team, so that each team's
performance can be better evaluated.

```{r pie_chart, message=FALSE, warning=FALSE}
# rendering the pie chart
# function returning the distribution of players's participation percentage,
# the players' GP over the total number of games in the season, of a selected team

# we are displaying two teams' player participation percentage distribution as
# examples, the team with the most number of players participated in more than 90%
# games and the team with the most number of players participated in less than 30%
# games of the season (displaying all team's chart will be too many for the report
# as we have 30 teams participated in this season)
san_player_GP_distribution <- get_pie_chart(regular_season_df, "San")
phi_player_GP_distribution <- get_pie_chart(regular_season_df, "Phi")
plot(san_player_GP_distribution)
plot(phi_player_GP_distribution)
```

### Important Insights from the Pie Chart
The first chart is for San Antonio Spurs, the team with the most number of players
participated in more than 90% games in the league. San finished the season with 7th
place at western conference. The second chart is for Philadelphia 76ers, the team
with the most number of players participated in no more than 30% games of the
season in the league, who finished the season with 3rd place at eastern conference.
In total, there were 82 games in the season. The pie chart shows the distribution
of players' participation rate which is defined as the player's GP over 82, the
total number of games in this season, of a selected team. It is important to
understand that the participation rate distribution does not necessaily define the
team's ability. A team with multiple players who participated in most of the games
means the team has more experienced players, but the team can still lose sometimes,
and similarly, the team with multiple players who only participated in small number
of games in the season can win sometimes as well. 
